An Empirical Analysis of the Role of Energy in Economic Growth

### 2013-02-25 EconGrowth.Rnw
Now including code to generate a graph of SE vs. gamma for the CD models. Trying to debug why the ZA model doesn't converge.

<<setup, echo=FALSE, message=FALSE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
require(micEconCES)
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=TRUE,
  # ********************************************************************************************************
  warning=TRUE, #Tells whether to show warnings in the output.
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (cache=TRUE).
  cache=FALSE,
  echo=FALSE,    #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
  message=FALSE) #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.

@@

<<cobb-douglas functions, eval=TRUE>>=
####################
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
##
cobbDouglasModel <- function(countryAbbrev, energyType){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  #Run the non-linear least squares fit to the data based on whether we have an energy type or not.
  if (missing(energyType)){
    # No energy term desired in the Cobb-Douglas equation.
    # Establish guess values for alpha and lambda.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.7 #0.7 gives good results for all countries.  
    start <- list(lambda=lambdaGuess, alpha=alphaGuess)
    # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
    model <- iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha)
    modelCD <- nls(formula=model, data=dataTable, start = start,
                   #Include the next 3 lines to fit with constraints.
                   #algorithm = "port",
                   #lower = list(lambda=-Inf, alpha=0), #constraints on lambda and alpha
                   #upper = list(lambda=Inf, alpha=1)
                   )
    return(modelCD)
  } else {
    # We need to do the Cobb-Douglas fit with the desired energy data.
    # To achieve the correct fit, we'll change the name of the desired column
    # to "iEToFit" and use "iEToFit" in the nls function. 
    dataTable <- replaceColName(dataTable, energyType, "iEToFit")
    # Reparameterize to ensure that we meet the constraints:
    # * alpha + beta + gamma = 1.0.
    # * alpha, beta, and gamma are all between 0.0 and 1.0.
    # To do this, we reparameterize as
    # * 0 < a < 1
    # * 0 < b < 1
    # * alpha = min(a, b)
    # * beta = b - a
    # * gamma = 1 - max(a, b)
    # Establish guess values for lambda, alpha, and beta.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.9
    betaGuess <- 1.0 - alphaGuess
    if (countryAbbrev == "ZA"){
      # South Africa doesn't converge using the default guess values above. Try different values.
      alphaGuess <- 0.597 # The alpha value when energy is not included
      betaGuess <- 1.0 - alphaGuess
#       alphaGuess <- 0.4
#       gammaGuess <- 0.1
#       betaGuess <- 1.0 - alphaGuess - gammaGuess
#       alphaGuess <- 0.39
#       betaGuess <- 0.41
    }
    aGuess <- alphaGuess
    bGuess <- alphaGuess + betaGuess
    start <- list(lambda=lambdaGuess, a=aGuess, b=bGuess)
    # Now actually do the fit, using the column name "iEToFit".
    model <- iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b))
    modelCDe <- nls(formula=model, data = dataTable, start = start,
                    control = nls.control(maxiter = 200, 
                                          tol = 1e-05, 
                                          minFactor = 1/1024, 
                                          printEval=FALSE, #Tells whether to print details of curve fit process.
                                          warnOnly=TRUE),
                    #Include the next 3 lines to fit with constraints.
                    #algorithm = "port",
                    #lower = list(lambda=-Inf, a=0, b=0),
                    #upper = list(lambda= Inf, a=1, b=1)
                    )
    return(modelCDe)
  }
}

>>>>

<<setup, echo=FALSE, message=FALSE>>=
require(knitr)
require(lattice)
require(latticeExtra)
require(ggplot2)
require(car)
require(mosaic)
require(xtable)
require(nlmrt)
require(micEconCES)
# tikz allows use of LaTeX formatting and font in graphs. Allows for a consistent look across the paper.
# See http://r-forge.r-project.org/R/?group_id=440 for instructions on installing tikzDevice.
require(tikzDevice) 
trellis.par.set(theme=theme.mosaic())
options(width=75)
options(show.signif.stars=TRUE)
opts_chunk$set(
  # ********************************************************************************************************
  eval=FALSE,
  # ********************************************************************************************************
  warning=TRUE, #Tells whether to show warnings in the output.
  dev='tikz',    #Allows LaTeX code in graphical output. E.g., "$y$" for a variable name in the legend for a graph.
  tidy=FALSE,
  comment=NA,
  #Tells whether to cache output from chunks are saved. Cacheing saves time. 
  #However, references to LaTeX tables DO NOT WORK if cacheing is turned on (cache=TRUE).
  cache=FALSE,
  echo=FALSE,    #Tells whether to echo code for all chunks. Set TRUE on an individual chunk to see its code.
  message=FALSE) #Turns off messages for all chunks. Set TRUE on an individual chunk to see it.

@@

<<cobb-douglas functions, eval=TRUE>>=
# Returns an nls Cobb-Douglas model for the country specified
# Give an energyType ("Q", "X", or "U") if you want to include an energy term
# by including the gamma parameter, a fit with fixed gamma will be provided.
cobbDouglasModel <- function(countryAbbrev, energyType, gamma){
  dataTable <- loadData(countryAbbrev) #Load the data that we need.
  #Run the non-linear least squares fit to the data based on whether we have an energy type or not.
    # No energy term desired in the Cobb-Douglas equation.
    # Establish guess values for alpha and lambda.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.7 #0.7 gives good results for all countries.  
    start <- list(lambda=lambdaGuess, alpha=alphaGuess)
    # Runs a non-linear least squares fit to the data. We've replaced beta with 1-alpha for simplicity.
    model <- iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha)
    modelCD <- nls(formula=model, data=dataTable, start = start,
                   #Include the next 3 lines to fit with constraints.
                   #algorithm = "port",
                   #lower = list(lambda=-Inf, alpha=0),
                   #upper = list(lambda=Inf, alpha=1)
    return(modelCD)
    # We need to do the Cobb-Douglas fit with the desired energy data.
    # To achieve the correct fit, we'll change the name of the desired column
    # to "iEToFit" and use "iEToFit" in the nls function. 
    dataTable <- replaceColName(dataTable, energyType, "iEToFit")
    # Reparameterize to ensure that we meet the constraints:
    # * alpha + beta + gamma = 1.0.
    # * alpha, beta, and gamma are all between 0.0 and 1.0.
    # To do this, we reparameterize as
    # * 0 < a < 1
    # * 0 < b < 1
    # * alpha = min(a, b)
    # * beta = b - a
    # * gamma = 1 - max(a, b)
    # Establish guess values for lambda, alpha, and beta.
    lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
    alphaGuess <- 0.9
    betaGuess <- 1.0 - alphaGuess
    if (countryAbbrev == "ZA"){
      # South Africa doesn't converge using the default guess values above. Try different values.
      alphaGuess <- 0.597 # The alpha value when energy is not included
      betaGuess <- 1.0 - alphaGuess
#       alphaGuess <- 0.4
#       gammaGuess <- 0.1
#       betaGuess <- 1.0 - alphaGuess - gammaGuess
#       alphaGuess <- 0.39
#       betaGuess <- 0.41
    aGuess <- alphaGuess
    bGuess <- alphaGuess + betaGuess
    start <- list(lambda=lambdaGuess, a=aGuess, b=bGuess)
    # Now actually do the fit, using the column name "iEToFit".
    if (missing(gamma)){
      # gamma is a free parameter that we want to estimate
      model <- iGDP ~ exp(lambda*iYear) * iCapStk^min(a,b) * iLabor^abs(b-a) * iEToFit^(1.0 - max(a,b))
      modelCDe <- nls(formula=model, data = dataTable, start = start,
                      control = nls.control(maxiter = 200, 
                                            tol = 1e-05, 
                                            minFactor = 1/1024, 
                                            printEval=FALSE, #Tells whether to print details of curve fit process.
                                            warnOnly=TRUE),
                      #Include the next 3 lines to fit with constraints.
                      #algorithm = "port",
                      #lower = list(lambda=-Inf, a=0, b=0),
                      #upper = list(lambda= Inf, a=1, b=1)
      # We don't want to estimat the value of gamma
      # Calculate a new column iEGamma = iEToFit^gamma
      iEGamma <- with(dataTable, iEToFit^gamma)
# print(iEGamma)
      dataTable <- cbind(dataTable, iEGamma)
# print(dataTable)
      # Now fit the CD model with fixed gamma
      # Establish guess values for alpha and lambda.
      lambdaGuess <- 0.0 #guessing lambda = 0 means there is no technological progress.
      alphaGuess <- 0.7 #0.7 gives good results for all countries.  
      start <- list(lambda=lambdaGuess, alpha=alphaGuess)
      model <- iGDP ~ exp(lambda*iYear) * iCapStk^alpha * iLabor^(1 - alpha) * iEGamma
      modelCDe <- nls(formula=model, data = dataTable, start = start,
                      control = nls.control(maxiter = 200, 
                                            tol = 1e-05, 
                                            minFactor = 1/1024, 
                                            printEval=FALSE, #Tells whether to print details of curve fit process.
                                            warnOnly=TRUE),
                      #Include the next 3 lines to fit with constraints.
                      #algorithm = "port",
                      #lower = list(lambda=-Inf, alpha=0),
                      #upper = list(lambda=Inf, alpha=1)
# print(summary(modelCDe))
    return(modelCDe)

###############################
# Does a grid search over values of gamma for a CD production function
##
cdGrid <- function(countryAbbrev, energyType, gammaGrid){
  gamma <- c(0.1, 0.2)
  gamma <- gammaGrid
  models <- lapply(gamma, cobbDouglasModel, energyType=energyType, countryAbbrev=countryAbbrev)
  return(models)
}

seFromNlsModel <- function(model) { 
  return(summary(model)$sigma)
}

gammaGrid <- seq(0.0, 1.0, by=0.1)
models <- cdGrid("US", "Q", gammaGrid)
SSE <- lapply(models, function(model) summary(model)$sigma)
data <- cbind(gammaGrid, SSE)
print("US")
plot(gammaGrid, SSE)
print(coef(models[[2]]))

models <- cdGrid("ZA", "Q", gammaGrid)
SSE <- lapply(models, function(model) summary(model)$sigma)
data <- cbind(gammaGrid, SSE)
print("ZA")
plot(gammaGrid, SSE)
print(coef(models[[5]]))
